---
title: 'Sankey visualization of protein-based synteny across chicken, zebra finch and new barn owl assemblies'
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Set up

Load libraries:

```{r echo=TRUE}
library(ggplot2)
library(httpgd)
library(dplyr)
library(stringr)
library(networkD3)

setwd('~/REMOTE_WORKING/refgenome2025/parsed_alignments/')
```

Since running *miniprot*, the father and mother chromosome names have been assigned. Load renaming tables:

```{r echo=TRUE}
nnamesFA <- read.table("C:/Users/mbachma3/Documents/REMOTE_WORKING/refgenome2025/2.1_CorrespondanceFather.txt") %>% select(V1, V6)
colnames(nnamesFA) <- c("tname", "ntname")
nnamesMO <- read.table("C:/Users/mbachma3/Documents/REMOTE_WORKING/refgenome2025/2.2_CorrespondanceMother.txt") %>% select(V1, V6)
colnames(nnamesMO) <- c("tname", "ntname")
```

## Father assembly

Load *miniprot* and feature annotation results:

```{r echo=TRUE}
chicken_barnFA <- read.table("sankey_chicken_to_barnowlFA.tsv", sep = '\t', header = TRUE)%>% 
  left_join(nnamesFA, by = "tname")
zfinch_barnFA <- read.table("sankey_zfinch_to_barnowlFA.tsv", sep = '\t', header = TRUE)%>% 
  left_join(nnamesFA, by = "tname")

```

The "chromosome" column keeps, for the chicken, the original names of the unplaced scaffolds. The "chr" column groups all these into a single "unplaced" category. The zebra finch assembly did not include unplaced scaffolds.


Count the number of times a protein in a chicken chromosome matches in another barn owl chromosome. The same is done for the zebra finch - barn owl matches. But to do a 3-way Sankey, set the source-target as follows:

|source |target  |
|-------|--------|
|chicken|b. owl  |
|b. owl |z. finch|


```{r}
linksGG <- chicken_barnFA %>%
    group_by(source = chromosome, target = ntname) %>%
    summarise(value = n(), .groups = "drop") %>%
    ungroup() %>%
  mutate(source = str_c(" ", source), 
         source = str_replace_all(source, "unplaced_", ""))

linksZF <- zfinch_barnFA %>%
    group_by(source = ntname, target = chromosome) %>%
    summarise(value = n(), .groups = "drop") %>%
    ungroup()
```

Check if there are any barn owl chromosomes not linked to either species. If it is the case, it is necessary to add placeholder links so that the nodes are not shifted and we keep chicken | barn owl | zebra finch order.

```{r}
nnamesFA %>% filter(! ntname %in% linksGG$target)
nnamesFA %>% filter(! ntname %in% linksZF$source)
``` 

Minimum two protein hits are required to link one chicken or zebra finch chromosome with a barn owl chromosome:

```{r}
allchromosomes <- rbind(linksGG, linksZF) %>% 
  filter(value >= 2)
```

```{r}
# prepare nodes data frame 
# create a sort key so the chromosomes can be ordered by name
nodesALL <- data.frame(name = unique(c(allchromosomes$source, allchromosomes$target))) %>%
  mutate(
    species_group = case_when(
      str_starts(name, " ")  ~ "1.chicken",
      str_starts(name, "Fa") ~ "2.barn_owl",
      TRUE                   ~ "3.zebra_finch"
    ),
    sort_key = case_when(
      name == " W"                        ~ 40,
      name == " Z"                        ~ 41,
      name == "W"                         ~ 40,
      name == "Z"                         ~ 41,
      name == "1A"                        ~ 1.5,
      name == "4A"                        ~ 4.5,
      name == "Fa_chrZ"                   ~ 46,
      str_detect(name, "\\d+")           ~ as.numeric(str_extract(name, "\\d+"))
    )
  ) %>%
  arrange(species_group, sort_key, name) %>%
  select(-sort_key)

# Convert to zero-based indices
allchromosomes$IDsource <- match(allchromosomes$source, nodesALL$name) - 1
allchromosomes$IDtarget <- match(allchromosomes$target, nodesALL$name) - 1

my_color <- 'd3.scaleOrdinal()
  .domain(["1.chicken", "2.barn_owl", "3.zebra_finch"])
  .range(["#808080", "#81A2A6", "#808080"])'

  
  # Create sankey diagram
sankeyNetwork(
  Links = allchromosomes,
  Nodes = nodesALL,
  Source = "IDsource",
  Target = "IDtarget",
  Value = "value",
  NodeID = "name",
  sinksRight = FALSE,
  fontSize = 16,
  nodeWidth = 30, 
  iterations = 0,
  height = 1200, 
  width = 2600, 
  nodePadding = 15,
  colourScale=my_color, NodeGroup="species_group"
  
  
) %>%
onRender("
  function(el, x) {
    var svg = d3.select(el).select('svg');
    var titles = ['Chicken', 'Barn owl (FA)', 'Zebra finch'];

    // Get the x positions of each column by reading node positions
    var nodePositions = [];
    svg.selectAll('.node').each(function(d) {
      var xPos = d.x;
      if (!nodePositions.includes(xPos)) {
        nodePositions.push(xPos);
      }
    });
    nodePositions.sort(function(a, b) { return a - b; });

    // Add a title above each column
    nodePositions.forEach(function(xPos, i) {
      svg.append('text')
        .attr('x', xPos)
        .attr('y', 15)
        .attr('text-anchor', 'middle')
        .attr('font-size', '20px')
        .attr('font-weight', 'bold')
        .attr('fill', '#333')
        .text(titles[i]);
    });
  }
")
```


## Mother assembly

Simlar steps as for father assembly:

```{r echo=TRUE}
chicken_barnMO <- read.table("sankey_chicken_to_barnowlMO.tsv", sep = '\t', header = TRUE)%>% 
  left_join(nnamesMO, by = "tname")
zfinch_barnMO <- read.table("sankey_zfinch_to_barnowlMO.tsv", sep = '\t', header = TRUE)%>% 
  left_join(nnamesMO, by = "tname")

```

```{r}
linksGG_mo <- chicken_barnMO %>%
    group_by(source = chromosome, target = ntname) %>%
    summarise(value = n(), .groups = "drop") %>%
    ungroup() %>%
  mutate(source = str_c(" ", source), 
         source = str_replace_all(source, "unplaced_", ""))

linksZF_mo <- zfinch_barnMO %>%
    group_by(source = ntname, target = chromosome) %>%
    summarise(value = n(), .groups = "drop") %>%
    ungroup()
```

Check if there are any barn owl chromosomes not linked to either species. If it is the case, it is necessary to add placeholder links so that the nodes are not shifted and we keep chicken | barn owl | zebra finch order.

```{r}
nnamesMO %>% filter(! ntname %in% linksGG_mo$target)
nnamesMO %>% filter(! ntname %in% linksZF_mo$source)
``` 

Minimum two protein hits are required to link one chicken or zebra finch chromosome with a barn owl chromosome:

```{r}
allchromosomes_mo <- rbind(linksGG_mo, linksZF_mo) %>% 
  filter(value >= 2)
```

```{r}
# prepare nodes data frame 
# create a sort key so the chromosomes can be ordered by name
nodesALL_mo <- data.frame(name = unique(c(allchromosomes_mo$source, allchromosomes_mo$target))) %>%
  mutate(
    species_group = case_when(
      str_starts(name, " ")  ~ "1.chicken",
      str_starts(name, "Mo") ~ "2.barn_owl",
      TRUE                   ~ "3.zebra_finch"
    ),
    sort_key = case_when(
      name == " W"                        ~ 40,
      name == " Z"                        ~ 41,
      name == "W"                         ~ 40,
      name == "Z"                         ~ 41,
      name == "1A"                        ~ 1.5,
      name == "4A"                        ~ 4.5,
      name == "Fa_chrZ"                   ~ 46,
      str_detect(name, "\\d+")           ~ as.numeric(str_extract(name, "\\d+"))
    )
  ) %>%
  arrange(species_group, sort_key, name) %>%
  select(-sort_key)

# Convert to zero-based indices
allchromosomes_mo$IDsource <- match(allchromosomes_mo$source, nodesALL_mo$name) - 1
allchromosomes_mo$IDtarget <- match(allchromosomes_mo$target, nodesALL_mo$name) - 1

mo_color <- 'd3.scaleOrdinal()
  .domain(["1.chicken", "2.barn_owl", "3.zebra_finch"])
  .range(["#808080", "#733B37", "#808080"])'

  
  # Create sankey diagram
sankeyNetwork(
  Links = allchromosomes_mo,
  Nodes = nodesALL_mo,
  Source = "IDsource",
  Target = "IDtarget",
  Value = "value",
  NodeID = "name",
  sinksRight = FALSE,
  fontSize = 16,
  nodeWidth = 30, 
  iterations = 0,
  height = 1200, 
  width = 2600, 
  nodePadding = 15,
  colourScale=mo_color, NodeGroup="species_group"
  
  
) %>%
onRender("
  function(el, x) {
    var svg = d3.select(el).select('svg');
    var titles = ['Chicken', 'Barn owl (MO)', 'Zebra finch'];

    // Get the x positions of each column by reading node positions
    var nodePositions = [];
    svg.selectAll('.node').each(function(d) {
      var xPos = d.x;
      if (!nodePositions.includes(xPos)) {
        nodePositions.push(xPos);
      }
    });
    nodePositions.sort(function(a, b) { return a - b; });

    // Add a title above each column
    nodePositions.forEach(function(xPos, i) {
      svg.append('text')
        .attr('x', xPos)
        .attr('y', 15)
        .attr('text-anchor', 'middle')
        .attr('font-size', '20px')
        .attr('font-weight', 'bold')
        .attr('fill', '#333')
        .text(titles[i]);
    });
  }
")
```